---
title: "CS 422 Section 01"
author: "Josh Bowden"
output: html_notebook
---

 **1.1.** By taking this course, I foremost hope to learn how to use machine learning to create models and make sense of larger datasets. I also hope to learn how to use R more since I do have some experience with the Python side of things. I'd also hope to get a better understanding of how to use fundamental statistics for understanding your data.


## Problem 1

 **1-A**: Load the *cars* dataset and print it.
```{r}
data(cars)
cars
```

 **1-B**: Plot the data in the *cars* dataset. Provide a title for the graph. The title should be "The Cars Dataset".
```{r}
plot(cars,
     main = "The Cars Dataset",
     xlab = "Speed (mph)",
     y = "Stopping Distance (ft)")
```

 **1-C**: Print a summary of the cars dataset using the R `summary()` command.
```{r}
summary(cars)
```

 **1-D**: What is the maximum speed and minimum distance as shown in the `summary()` command?
 
> The minimum distance is 2ft while the maximum distance is 120ft.


## Problem 2

 **2-A**: Read the file into a data frame and add a new row consisting of the following data to the existing rows:

> name: Brad Pitt  
  id: 40051  
  gpa: 2.21  
  grade: C  


Before adding the student:

```{r}
students <- read.csv("student.csv", sep=";", header=T)
students
```


After adding the new student:

```{r}
brad_pitt <- data.frame(name = "Brad Pitt",
                        id = 40051,
                        gpa = 2.21,
                        grade = "C")
students.new <- rbind(students, brad_pitt)
students.new
```

